---
title: "Getting Started with EvalEmulateSurv"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with EvalEmulateSurv}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  warning   = FALSE,
  message   = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Introduction

## Background and Purpose

Individual Patient Data (IPD) meta-analysis is considered the gold standard
for evidence synthesis in clinical research. However, access to original IPD
is often restricted due to data privacy regulations and proprietary
considerations. To circumvent this barrier, data emulation methods have been
developed that reconstruct plausible synthetic IPD from published
Kaplan--Meier (KM) curves and summary statistics.

Once emulated IPD has been generated, a critical question arises: **how
faithfully does the emulated data reproduce the statistical properties of the
original IPD?** The **EvalEmulateSurv** package provides a set of functions
designed to answer this question quantitatively, using three complementary
metrics:

1. **NAUC** (Normalised Area Under the Curve)
2. **K-S distance** (Kolmogorov--Smirnov distance)
3. **RMSE** (Root Mean Square Error)

All three metrics compare the KM survival curves estimated from the original
IPD and the emulated IPD, with smaller values indicating closer agreement. A
key design principle of this package is that **all KM estimation is
implemented from scratch** using base R, without relying on the `survival`
package, making the functions portable and dependency-light.

## Package Setup

```{r}
library(EvalEmulateSurv)
library(ggplot2)
```

## Data Preparation

Throughout this vignette, we use dummy data generated by `gen_dummy_data()`.
All functions in this package assume that `data_group1` (original IPD) and
`data_group2` (emulated IPD) contain **a single survival endpoint**. When
your dataset contains multiple endpoints (e.g., OS and PFS in the same
data frame), filter to one endpoint before passing to any function.

```{r}
# Generate dummy IPD
ipd <- gen_dummy_data(seed = 42)

# Filter to OS endpoint and split by group
ipd_os  <- ipd[ipd$TYPE == "OS", ]
grp1_os <- ipd_os[ipd_os$GROUP == "1", ]   # group 1 (original IPD)
grp2_os <- ipd_os[ipd_os$GROUP == "2", ]   # group 2 (emulated IPD)

# Truncation time for NAUC and K-S distance
tau_os <- 36
```

The dataset contains the following variables.

```{r}
head(grp1_os)
```

---

# Core Function: `km_est()`

## Purpose

`km_est()` is the foundational function of the package. It computes KM step
functions for two groups from raw IPD and serves as the common input for all
downstream metric functions. It also computes Greenwood variance, pointwise
confidence bands, and descriptive statistics (sample size, event count,
median survival time with confidence interval) for each group.

## Statistical Background

### Kaplan--Meier Estimator

The Kaplan--Meier estimator at each event time $t_{(j)}$ is:

$$\hat{S}(t_{(j)}) = \prod_{i:\, t_{(i)} \le t_{(j)}}
\left(1 - \frac{d_i}{n_i}\right)$$

where $d_i$ is the number of events and $n_i$ is the number at risk at
time $t_{(i)}$. The resulting step function is constant on the half-open
interval $[t_{(j)},\, t_{(j+1)})$ and drops at each event time.

### Greenwood Variance and Confidence Band

The variance of $\hat{S}(t)$ is estimated by Greenwood's formula:

$$\widehat{\mathrm{Var}}(\hat{S}(t)) =
\hat{S}(t)^2 \sum_{t_{(j)} \le t} \frac{d_j}{n_j(n_j - d_j)}$$

A pointwise 95% confidence band is constructed on the log scale:

$$\exp\!\left(\log \hat{S}(t) \pm
\frac{1.96\,\sqrt{\widehat{\mathrm{Var}}(\hat{S}(t))}}{\hat{S}(t)}\right)$$

This is equivalent to `survival::survfit` with `conf.type = "log"`.

### Median Survival Time and CI

The median is the smallest $t$ such that $\hat{S}(t) \le 0.5$.
The confidence interval for the median is derived from the pointwise
confidence band: the lower bound is the smallest $t$ where the lower
confidence band $\le 0.5$, and the upper bound is the smallest $t$ where
the upper confidence band $\le 0.5$.

## Usage

```{r}
km_os <- km_est(
  data_group1 = grp1_os,
  data_group2 = grp2_os,
  time_var    = "TIME",
  event_var   = "EVENT",
  tau         = tau_os
)

# KM step function for group 1 (time, surv, var_surv, surv_lo, surv_hi)
head(km_os$km1)

# Descriptive statistics for group 1
km_os$desc1

# Descriptive statistics for group 2
km_os$desc2

# Truncation time and maximum observation time
km_os$tau
km_os$t_max_obs
```

## Validation Against the `survival` Package

The implementation is numerically equivalent to `survival::survfit`. The
maximum absolute difference in survival estimates, confidence band bounds,
and median survival time is within floating-point machine epsilon
($\varepsilon_{\mathrm{machine}} \approx$
`r format(.Machine$double.eps, scientific = TRUE)`), confirming numerical
equivalence.

---

# Metric 1: NAUC via `nauc()`

## Statistical Definition

The **Normalised Area Under the Curve (NAUC)** measures the normalised
absolute area between two KM survival curves over $[0, \tau]$:

$$\widehat{\mathrm{NAUC}}(\tilde{d}_n \mid d_n) :=
\frac{1}{\tau} \int_0^{\tau} \left| \hat{S}(t) - \tilde{S}(t) \right| dt
\in [0, 1]$$

where $\hat{S}(t)$ is the KM estimate from the original IPD (group 1,
ground truth) and $\tilde{S}(t)$ is the KM estimate from the emulated IPD
(group 2).

### Interpretation

NAUC measures the **mean absolute vertical distance** between two KM curves
over $[0, \tau]$. Geometrically, $\tau \times \mathrm{NAUC}$ equals the
total grey shaded area in the KM plot produced by `kmplot()`.

**Note:** NAUC differs from the absolute RMST difference. The RMST
difference uses the signed integral, allowing cancellation when curves cross.
NAUC uses the absolute value, so cancellation cannot occur:

$$\mathrm{NAUC} \ge \frac{1}{\tau}
\left|\int_0^\tau \left[\hat{S}(t) - \tilde{S}(t)\right] dt\right|
= \frac{|\mathrm{RMST}_1(\tau) - \mathrm{RMST}_2(\tau)|}{\tau}$$

### Discrete Estimator

Let the distinct ordered event times up to $\tau$ across both groups be
$0 = t_{(0)} \le \cdots \le t_{(q)} \le \tau$ with
$\delta_j = t_{(j)} - t_{(j-1)}$. The discrete estimator is:

$$\widehat{\mathrm{NAUC}} =
\frac{1}{\tau} \sum_{j=1}^{q} \delta_j
\left| \hat{S}(t_{(j)}) - \tilde{S}(t_{(j)}) \right|$$

### Subgroup Aggregation

When subgroups are specified, per-stratum NAUCs are aggregated using a
**sample-size-weighted average** (default, per Zhao et al. 2025):

$$\widehat{\mathrm{NAUC}}_{\mathrm{wtd}} =
\sum_{x=1}^{K} \frac{n_x}{n} \widehat{\mathrm{NAUC}}_x$$

A simple (unweighted) average is also available via
`method_avg = "simple"`.

**Note:** The aggregate NAUC is not equal to the NAUC computed directly on
the overall population. The overall NAUC uses the KM curves of the full
mixed population, whereas the aggregate is a weighted average of per-stratum
discrepancies. Due to the mixing of subpopulations, the overall KM curves may
be smoother and less discrepant than individual subgroup curves, so the
overall NAUC can be smaller than any per-stratum NAUC (analogous to
Simpson's paradox in a survival analysis context).

## Usage

```{r}
# Overall population
res_nauc_overall <- nauc(
  data_group1 = grp1_os,
  data_group2 = grp2_os,
  time_var    = "TIME",
  event_var   = "EVENT",
  tau         = tau_os
)

res_nauc_overall$by_stratum
res_nauc_overall$aggregate
```

```{r}
# Subgroup by REGION: weighted average (default)
res_nauc_region <- nauc(
  data_group1  = grp1_os,
  data_group2  = grp2_os,
  time_var     = "TIME",
  event_var    = "EVENT",
  tau          = tau_os,
  subgroup_var = "REGION",
  method_avg   = "weighted"
)

res_nauc_region$by_stratum
res_nauc_region$aggregate
```

```{r}
# Two-way interaction: SEX x REGION (2 x 3 = 6 strata)
res_nauc_inter <- nauc(
  data_group1  = grp1_os,
  data_group2  = grp2_os,
  time_var     = "TIME",
  event_var    = "EVENT",
  tau          = tau_os,
  subgroup_var = "SEX:REGION"
)

res_nauc_inter$by_stratum
```

---

# Metric 2: K-S Distance via `ksdist()`

## Statistical Definition

The **Kolmogorov--Smirnov (K-S) distance** is the supremum of the absolute
difference between two KM survival curves over $[0, \tau]$:

$$\widehat{KS}(\tilde{d}_n \mid d_n) :=
\sup_{0 \le t \le \tau} \left| \hat{S}(t) - \tilde{S}(t) \right|
\in [0, 1]$$

### Interpretation

The K-S distance measures the **single largest vertical gap** between the
two KM curves over $[0, \tau]$. It is a worst-case measure: while NAUC
captures the average discrepancy, K-S captures the maximum discrepancy at
any single time point. In `kmplot()`, it corresponds to the height of the
tallest grey rectangle, and the time at which this maximum is achieved is
indicated by the vertical dashed line.

### Discrete Estimator

Since both KM curves are step functions, the supremum reduces to a maximum
over the union of event times from both groups (excluding $t = 0$ where
both curves equal 1):

$$\widehat{KS} =
\max_{1 \le j \le q} \left| \hat{S}(t_{(j)}) - \tilde{S}(t_{(j)}) \right|$$

The time point $t_{\max\mathrm{gap}}$ at which the maximum is achieved is
also returned, useful for plot annotation.

### Behaviour When `tau` Exceeds One Group's Maximum Observation Time

When `tau` is set to a value that exceeds the maximum observed time in one
group (e.g., group 2's last observation is at $t = 30$ but `tau = 40`), the
KM step function of that group **carries forward the last observed survival
probability** for all $t > 30$. This is the standard convention for KM
curves: $\hat{S}(t)$ is defined as the last estimated value for $t$ beyond
the last event time. As a result, the K-S distance (and NAUC) for
$t \in (30, 40]$ is computed using this carry-forward value for group 2,
which is handled automatically by `findInterval()` inside `km_est()`.

### Subgroup Aggregation

Per Zhao et al. (2025), the K-S distance is aggregated using a **simple
(unweighted) average** (default):

$$\widehat{KS}_{\mathrm{avg}} = \frac{1}{K} \sum_{x=1}^{K} \widehat{KS}_x$$

Equal weighting reflects the fact that K-S distance measures the worst-case
scenario, and each stratum is considered equally important regardless of
sample size.

## Usage

```{r}
# Overall population
res_ks_overall <- ksdist(
  data_group1 = grp1_os,
  data_group2 = grp2_os,
  time_var    = "TIME",
  event_var   = "EVENT",
  tau         = tau_os
)

res_ks_overall$by_stratum   # includes t_max_gap
res_ks_overall$aggregate
```

```{r}
# Subgroup by REGION: simple average (default)
res_ks_region <- ksdist(
  data_group1  = grp1_os,
  data_group2  = grp2_os,
  time_var     = "TIME",
  event_var    = "EVENT",
  tau          = tau_os,
  subgroup_var = "REGION"
)

res_ks_region$by_stratum
res_ks_region$aggregate
```

---

# Metric 3: RMSE via `rmse()`

## Statistical Definition

The **Root Mean Square Error (RMSE)** is:

$$\mathrm{RMSE} = \sqrt{\frac{1}{M} \sum_{k=1}^{M}
\left( \hat{S}(t_k) - \tilde{S}(t_k) \right)^2}$$

where $t_1, \ldots, t_M$ are $M = 100$ evenly spaced time points from $0$
to $t_{\max}$ (the maximum observed time across both groups),
$\hat{S}(t_k)$ is the KM estimate from the original IPD, and
$\tilde{S}(t_k)$ is the KM estimate from the emulated IPD.

### Interpretation

Like NAUC, RMSE measures the overall discrepancy between two KM curves. The
key difference is that RMSE uses **squared differences**, giving
disproportionately greater weight to large discrepancies.

### Comparison of NAUC and RMSE

| Property | NAUC | RMSE |
|---|---|---|
| Evaluation grid | Union of event times (data-driven) | $M$ evenly spaced points (fixed) |
| Aggregation of differences | Absolute value $\times$ interval length | Squared $\times$ equal weight |
| Sensitivity to large outliers | Moderate | High (due to squaring) |
| Truncation time | Required ($\tau$) | Full range ($t_{\max}$) |
| Range | $[0, 1]$ | $[0, \infty)$ |

### Treatment of Truncation Time

Following Lang et al. (2025), RMSE is computed over the **full observation
range** by default (`tau = NULL`). When evaluation points $t_k$ exceed the
last event time of one group, the KM step function carries forward the last
observed survival probability. This is consistent with the standard
convention for KM curves.

## Usage

```{r}
# Overall population (full range, tau = NULL by default)
res_rmse_overall <- rmse(
  data_group1 = grp1_os,
  data_group2 = grp2_os,
  time_var    = "TIME",
  event_var   = "EVENT"
)

res_rmse_overall$by_stratum
res_rmse_overall$n_points   # = 100
```

```{r}
# Subgroup by REGION, weighted average
res_rmse_region <- rmse(
  data_group1  = grp1_os,
  data_group2  = grp2_os,
  time_var     = "TIME",
  event_var    = "EVENT",
  subgroup_var = "REGION",
  method_avg   = "weighted"
)

res_rmse_region$by_stratum
res_rmse_region$aggregate
```

---

# Visualisation: `kmplot()`

## Purpose

`kmplot()` produces KM plots comparing two groups, with the following
visual elements per panel:

- **KM step curves** for both groups over the full observation range
- **Grey shaded rectangles** between the two curves within $[0, \tau]$,
  representing $\tau \times \mathrm{NAUC}$
- **Vertical dashed line** at the time of maximum K-S gap ($t_{\max\mathrm{gap}}$)
- **Annotation block** in the upper-right corner showing $\tau$, NAUC,
  K-S distance, and RMSE

## Relationship Between Visual Elements and Metrics

| Visual element | Corresponding metric |
|---|---|
| Total grey shaded area | $\tau \times \mathrm{NAUC}$ |
| Grey shaded area $\div\, \tau$ | NAUC (annotated value) |
| Height of the tallest grey rectangle | K-S distance (annotated value) |
| Position of vertical dashed line | $t_{\max\mathrm{gap}}$ (time of max K-S gap) |
| RMSE in annotation | Computed over full observation range |

## Overall Population

```{r, fig.width = 7, fig.height = 5, out.width = "100%"}
plots_os <- kmplot(
  data_group1  = grp1_os,
  data_group2  = grp2_os,
  time_var     = "TIME",
  event_var    = "EVENT",
  tau          = tau_os,
  label_group1 = "Original",
  label_group2 = "Emulated"
)
plots_os[["Overall"]]
```

## Subgroup Panels

When `subgroup_var` is specified, one panel is produced per stratum using
`facet_wrap()`. The x-axis is unified across all panels to allow visual
comparison.

```{r, fig.width = 8, fig.height = 6, out.width = "100%"}
plots_region <- kmplot(
  data_group1     = grp1_os,
  data_group2     = grp2_os,
  time_var        = "TIME",
  event_var       = "EVENT",
  tau             = tau_os,
  subgroup_var    = "REGION",
  include_overall = TRUE,
  ncol            = 2,
  label_group1    = "Original",
  label_group2    = "Emulated"
)
plots_region[["REGION"]]
```

## Interaction Term Panels

N-way interaction terms are supported using `":"` as a separator:

```{r, fig.width = 8, fig.height = 8, out.width = "100%"}
plots_inter <- kmplot(
  data_group1  = grp1_os,
  data_group2  = grp2_os,
  time_var     = "TIME",
  event_var    = "EVENT",
  tau          = tau_os,
  subgroup_var = "SEX:REGION",
  ncol         = 2,
  label_group1 = "Original",
  label_group2 = "Emulated"
)
plots_inter[["SEX:REGION"]]
```

---

# Summary Table: `summarytable()`

## Purpose

`summarytable()` provides a comprehensive summary combining:

- **Descriptive statistics** per group: sample size, event count, median
  survival time, and its confidence interval
- **All three evaluation metrics**: NAUC, K-S distance, and RMSE

for the overall population and each level of one or more subgroup variables
in a single call.

## Notes on Aggregate Values

- `nauc_aggregate`, `ks_aggregate`, `rmse_aggregate` are `NA` for the
  Overall block (single stratum; aggregation is meaningless).
- For subgroup blocks, all strata within the same specification share the
  same aggregate value.
- NAUC uses a **weighted average** by default; K-S distance and RMSE use
  a **simple average** by default.
- RMSE is always computed over the full observation range (`tau = NULL`).

## Usage

```{r}
res <- summarytable(
  data_group1   = grp1_os,
  data_group2   = grp2_os,
  time_var      = "TIME",
  event_var     = "EVENT",
  tau           = tau_os,
  subgroup_vars = c("SEX", "REGION"),
  label_group1  = "Original",
  label_group2  = "Emulated"
)
```

Plain-text output (default):

```{r}
print(res)
```

Markdown table output (useful in R Markdown / Quarto documents):

```{r}
print(res, kable = TRUE, kable_format = "pipe")
```

## Subgroup and Interaction Term Specification

The `subgroup_vars` argument supports single column names, interaction
terms using `":"`, and mixed specifications in the same call:

```{r}
res_mixed <- summarytable(
  data_group1   = grp1_os,
  data_group2   = grp2_os,
  time_var      = "TIME",
  event_var     = "EVENT",
  tau           = tau_os,
  subgroup_vars = c("SEX", "REGION", "SEX:REGION"),
  label_group1  = "Original",
  label_group2  = "Emulated"
)
print(res_mixed, digits = 3)
```

---

# Input Data Formats

All functions accept `data_group1` and `data_group2` in the following
formats, handled internally by `load_ipd()`:

| Format | Example |
|--------|---------|
| `data.frame` or tibble in memory | `grp1_os` |
| Path to a `.csv` file | `"path/to/group1.csv"` |
| Path to a single-object `.RData` / `.rda` file | `"path/to/group1.RData"` |

For `.RData` files containing multiple objects, use `load_ipd()` explicitly:

```{r, eval = FALSE}
d1 <- load_ipd("both_groups.RData", rdata_object = "grp1_os")
d2 <- load_ipd("both_groups.RData", rdata_object = "grp2_os")
res <- summarytable(data_group1 = d1, data_group2 = d2, ...)
```

---

# Metric Benchmarks

The following benchmarks from the published literature serve as reference
values when interpreting metric results.

| Metric | Benchmark | Source |
|--------|-----------|--------|
| NAUC | $< 2.3 \times 10^{-2}$ | Zhao et al. (2025) |
| K-S distance | No explicit threshold | -- |
| RMSE | $0.002$--$0.009$ | Lang et al. (2025) |

Note that these benchmarks were derived from specific simulation settings and
should be interpreted with caution. Smaller values indicate better
reproduction of the original IPD by the emulated data.

---

# References

- Zhao, Z. et al. (2025). SynthIPD: Generating synthetic individual patient
  data from published Kaplan--Meier curves and summary statistics.
  *arXiv:2509.16466*.

- Lang, et al. (2025). RESOLVE-IPD: Reconstruction of Survival data with
  Latent Outcomes from published Evidence for Individual Patient Data
  meta-analysis.
  *arXiv:2511.01785*.

- Kaplan, E. L. and Meier, P. (1958). Nonparametric estimation from
  incomplete observations. *Journal of the American Statistical Association*,
  53(282), 457--481.
